Improving Multi-Domain Neural Machine Translation with Differential Data Selection

### 2021-04-12 2021/emnlp/mdmt.tex
start reshaping text

\newcommand{\fyTodo}[1]{\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyTodostar}[1]{\Todo*[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyDone}[1]{\done[FY]\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyFuture}[1]{\done[FY]\Todo[FY:]{\textcolor{red}{#1}}}
\newcommand{\fyDonestar}[1]{\done[FY]\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\revision}[1]{\textcolor{red}{#1}}
\newcommand{\revisiondel}[1]{}
\newcommand{\src}{\ensuremath{\mathbf{f}}} % source sentence
\newcommand{\trg}{\ensuremath{\mathbf{e}}} % target sentence
\newcommand{\domain}[1]{\texttt{\textsc{#1}}}
\newcommand{\system}[1]{\texttt{{#1}}}
\newcommand{\vlambda}{\ensuremath{\boldsymbol\lambda}\xspace} % parameters vector for a distribution
\newcommand{\indic}[1]{\ensuremath{\mathbb{I}(#1)}}
% \newcommand{\SB}[1]{\textcolor{green}{#1}}
% \newcommand{\SW}[1]{\textcolor{red}{#1}}
\newcommand{\SB}[1]{\textbf{#1}}
\newcommand{\SW}[1]{\underline{#1}}
% limits underneath
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\renewcommand\textfraction{.1}
\renewcommand\floatpagefraction{.95}
\newcommand{\sbcl}[2]{{\scriptsize #1 \hfill $|$ \hfill  #2}}
\usepackage{multirow}
\usepackage{adjustbox}

@@

\begin{document}
\maketitle
\begin{abstract}
  When building machine translation systems, one often needs to make the best out of heterogeneous sets of parallel data in training to achieve the best test performance in one or several domain(s) of interest. This multi-source (multi-domain) adaptation problem is often approached with instance selection or reweighting strategies, most of which pre-suppose an ad-hoc assessment of the relevance of training instances. In this paper, we study the recently proposed Differential Data Selection (DDS) model and explore its ability to serve as a generic framework for these various situations. Our experiments study both domain adaptation and multi-domain learning and show that DDS often enables to outperform more heuristic adaptation strategies. We also introduce a variant that boosts DDS performance of adapter-based multi-domain systems.\fyTodo{Keep this part ?}
%   This is a well-known issue, which has given raise to a rich litterature in under the umbrellas of domain adaptation or multi-domain

@@

\section{Introduction}\label{sec:intro}
A typical setting in machine translation (MT) is to collect the largest possible collection of parallel data for the chosen language pair, with the intent to achieve optimal performance for a task of interest. In such situations, the training data distribution is opportunistic, while the test data distribution is chosen and fixed; a key component in training is then to mitigate the detrimental effects of a possible mismatch between these distributions. Single-source and multi-source domain adaptation (DA) is a well-studied instance of this setting (see \cite{Chu2017comparison} for a review) , and so is multi-domain (MD) learning \cite{Chu18multilingual,Zeng18multidomain,Jiang19multidomain,Pham21revisiting}. A related situation is multi-lingual machine translation \cite{Firat16multiway,Ha16towards,Johnson17google,Arivazhagan19massively}\fyTodo{Add more recent work}, where the heterogeneity of training data not only corresponds to variations in topic, genre or register, but also in language.

@@

It is usual in MT to opportunistically collect training samples from several domains, which means that training instances are distributed according to the mixture $\mathcal{D}^s$ such that $\mathcal{D}^s(x) = \sum_{d=1}^{n_d} \lambda^{s}(d) \mathcal{D}_d(x)$, where $\{\lambda^{s}(d), d=1 \dots n_d\}$ are the corresponding mixture weights such that $\sum_d \lambda^{s}(d)=1$).

The main challenge is then to make the best of this heterogeneous data, with the aim to achieve the optimal performance for the intended test conditions. These might correspond to data from just one of the training domains, as in standard supervised (multi-source) domain adaptation; a more difficult case is when the test data is from one domain unseen in training (unsupervised domain adaptation); in multi-domain adaptation finally the test distribution is itself a mixture of domains, some of which may also be observed in training.  Without loss of generality, one may then assume that the test distribution takes the form $\mathcal{D}^{t}(x) = \sum_d \lambda^{t}(d) \mathcal{D}_d(x)$ - with only one non-null  $\lambda^{t}(d)$ in the case of domain adaptation.
These situations are illustrated in Figure~\ref{fig:mdmt-lambdas}.
\begin{figure}[h]
  \centering
  \includegraphics[width=0.48\textwidth]{mdmt-lambdas}
  \caption{Training and testing with distribution mismatch. We consider just three domains, and represent vectors of mixture weights $\lambda^{s}$ and $\lambda^{t}$ in the 3-dimensional simplex. Training with weights in (a) and testing with weights in (c) is standard multi-source domain adaptation to $d_2$, while (b)-(c) is unsupervised adaptation; training with weights in (a) and testing with weights in (d) is multi-domain learning, also illustrated with configurations (a)-(e) (training domain $d_1$ is not seen in test), and (b)-(d)  (test domain $d_2$ is unseen in training).}\label{fig:mdmt-lambdas}
\end{figure}

@@

A more practical standard approach to supervised DA is \emph{fine-tuning} \cite{Luong15stanford,Freitag16fast}, where $\lambda^{l}$ is allowed to vary during the course of learning. With our notations, this approach amounts to first learning an initial parameter value with all the data ($\forall d, \lambda^{l}(d) = \lambda^{s}(d)$), then to continue training with only batches from the test domain $d_t$ ($\lambda^{l}(d) = \indic{d = d_t}$) with $\indic{A}$ the indicator function for predicate $A$. Note that this strategy is potentially suboptimal, as some out-of-domain samples may contribute to the final performance due to eg.\ domain overlap. Optimizing the learning distribution in multi-domain settings is even more challenging, as the learner has to best take advantage of potential domains overlaps, and also of the fact that some domains are easier to learn than others.\fyTodo{How to measure this?}

Differentiable data selection (DDS), proposed by \citet{Wang20optimizing} and later applied to multilingual machine translation by \citet{Wang20balancing}, introduces an additional degree of freedom and allows $\lambda^{l}(d)$ to change multiple times in the course of training, as we now explain.

@@

\section{Differential Data Selection} \label{sec:dds}
\subsection{Basic principles}
Assume training data from $n_d$ domains $d_1 \dots d_{n_d}$, we denote the size of the training corpus from domain $d$ as  $N^{s}_d$, and $N^{s} = \sum_d N^{s}_d$ is the total number of training samples. We use $\widehat{\mathcal{D}^l_d}$ and $\widehat{\mathcal{D}^t _d}$ to denote the empirical train and test distributions for domain $d$, and \revision{$\widehat{\mathcal{D}}(x;\lambda) = \sum_{d} \lambda(d) \widehat{\mathcal{D}^l_d}(x)$} \fyTodo{$\widehat{\mathcal{D}}(x;\lambda) = \sum_{d} \lambda(d) \widehat{\mathcal{D}_d}(x)$}. In our setting,  $\lambda^t$, and hence $\widehat{\mathcal{D}^t}(x;\lambda^t)$ are fixed and predefined, approximated with an equivalent number of development corpora. Training optimizes the parameters $\theta$ of a neural translation model.

Sampling training domains using a distribution $\lambda^{l}(d)$ results in a parameter estimate $\theta(\lambda^{l})$, delivering performance that can be estimated on the development set using $\lambda^{t}$. \cite{Wang20optimizing,Wang20balancing} then formulate the problem of finding an optimal sampling strategy $\lambda^{l}_{*}$ maximizing performance on the development data as the following bi-level optimization problem:
\begin{equation} \label{eq:bilevel}
\begin{split}
\lambda^{l}_* &= \displaystyle{\mathop{\argmin}_{\lambda^{l}} J(\theta^*(\lambda^{l}), \lambda^{t})} \\
					& \theta^*(\lambda^{l}) = \displaystyle{\mathop{\argmin_{\theta}} J(\theta, \lambda^{l})}, 
\end{split}
\end{equation}
where we use the following notations:% \fyTodo{Overload Q, from domain to instances to pairs}

\begin{align*}
J(\theta, \lambda^{t} ) &= \displaystyle{\mathop{\sum}_{(x,y) \sim \widehat{\mathcal{D}^{t}(x)}} \ell(\theta,x,y)} \\
J(\theta, \lambda^{l}) &= \displaystyle{\mathop{E_{(x,y) \sim \widehat{\mathcal{D}^{l}(x)}}[\ell(\theta,x,y)]}} \\
l(\theta,x,y) &= - log P(\theta,x,y)
\end{align*}
Problem~\eqref{eq:bilevel} can be approached by coordinate gradient descent alternating between $\lambda^{l}$ and $\theta$. \citet{Wang20balancing} limit the search space of $\lambda^{l}$ to a parameterized differentiable function $\lambda^{l}(d; \psi)$. Given $\psi_t$ at time step $t$, $\theta_t$ is updated via usual gradient descent as follows:
\begin{align*}
\theta_{t+1} &= \theta_t - \mathbf{lr}_{nmt} * \frac{\partial l(\theta_t, x,y)}{\partial \theta} \\
x,y &\sim \widehat{\mathcal{D}^l}(x;\lambda^l(\psi_t)) = \displaystyle{\mathop{\sum}_{d=1}^K} \lambda^l(d,\psi_t) * \widehat{D_d}(x) \\
\end{align*}
Given $\theta_t$, $\psi_t$ is updated via the Reinforce algorithm \cite{Williams92reinforce}:\fyTodo{Rewrite this correctly with domain average reward}
%\begin{align*}
%  \psi_{t+1} &= \psi_t + \mathbf{lr}_{data} * R(x,y) * \\
%  & \quad\quad \frac{\partial log(\lambda^l(\psi_t)(x,y))}{\partial \psi} \\
%x,y &\sim \lambda^l (\psi_t) \\
%R(x,y) &= \langle \frac{\partial l(\theta',x,y)}{\partial \theta}, \frac{\partial J(\theta_{t+1}, \lambda^l)}{\partial \theta} \rangle \\
%\theta' &= \theta_t - \mathbf{lr}_{nmt} * \frac{\partial l(\theta_t, x,y)}{\partial \theta} \\
%\end{align*}

@@

This setting is pretty general and can in principle accomodate the variety of situations mentioned  above, and many more: basic domain adaption, multi-domain adaptation with various target distributions, possibly including domains unseen in training. In our experiments we would like to better assess the actual potential of DDS is these settings, and seek to experimentally answer the following questions:
\begin{itemize}
\item is DDS a viable alternative to conventional fine-tuning ? In particular does it enable to better take advantage of relevant data from other domains ?
\item is DDS a viable alternative in multidomain scenarios ?
\item \fyTodo{TBContinued}
\end{itemize}

@@

\subsection{Data and metrics \label{ssec:corpora}}
We experiment with translation from English into French and use texts initially originating from 6~domains, corresponding to the following data sources: the UFAL Medical corpus V1.0 (\domain{med})\footnote{\url{https://ufal.mff.cuni.cz/ufal_medical_corpus}. \revision{We only use the in-domain (medical) subcorpora: PATR, EMEA, CESTA, ECDC.}}, the European Central Bank corpus (\domain{bank}) \cite{Tiedemann12parallel}; The JRC-Acquis Communautaire corpus (\domain{law}) \cite{Steinberger06acquis}, documentations for KDE, Ubuntu, GNOME and PHP from Opus collection \cite{Tiedemann09news}, collectively merged in a \domain{it}-domain, Ted Talks (\domain{talk}) \cite{Cettolo12wit}, and the Koran (\domain{rel}). Complementary experiments also use v12 of the News Commentary corpus (\domain{news}). Most corpora are available from the Opus web site.\footnote{\url{http://opus.nlpl.eu}} These corpora were deduplicated and tokenized with in-house tools; statistics are in Table~\ref{tab:Corpora}. To reduce the number of types and build open-vocabulary systems, we use Byte-Pair Encoding \cite{Sennrich16BPE} with 30,000 merge operations on a corpus containing all sentences in both languages.\fyDone{Add \# number of tokens, also specificity ?}%

@@

\begin{itemize}
\itemsep0em 
\item Generic models trained with various predefined mixtures of the training data taking the form:
\begin{align} \label{mixture:trn}
\lambda^{l}(d;\alpha) = \frac{q_d^{\alpha}}{\displaystyle{\mathop{\sum}_{d=1}^{n_d}q_d^{\alpha}}} &&
q_d = \frac{\mid N^{s}_d \mid}{\displaystyle{N^{s}}} % \mathop{\sum}_{i=1}^K\mid D_i \mid}}
\end{align} 
with $\alpha \in [0,0.25,0.5,0.75,1.0]$. These systems are denoted \system{Mixed-$\alpha$} below.
%on a concatenation of all corpora (\texttt{Mixed}). We develop two versions\footnote{In fact three: to enable a fair comparison with WDCMT, a RNN-based variant is also trained and evaluated. \revision{This system appears as \system{Mixed-Nat-RNN} in Table~\ref{tab:performance}}.} of this system, one where the domain unbalance reflects the distribution of our training data \revision{given in Table~\ref{tab:Corpora}} (\system{Mixed-Nat}) and one where all domains are equally represented in training (\system{Mixed-Bal}). The former is the best option when the train mixture $\mathcal{D}^s$ is also expected in testing; the latter should be used when the test distribution is uniform across domains. Accordingly, we report two aggregate scores: a weighted average reflecting the training distribution, and an unweighted average, meaning that test domains are equally important.
\item fine-tuned models \cite{Luong15stanford,Freitag16fast}, based on the \system{Mixed-$1.0$} system, further trained on each domain for at most 20~000 iterations, with early stopping when the dev BLEU stops increasing. The full fine-tuning (\system{FT-Full}) procedure may update all the parameters of the initial generic model, resulting in six systems each adapted for one domain, with no parameter sharing across domains.

@@

All models use embeddings and the hidden layers sizes of dimension~512. Transformer models contain with 8 attention heads in each of the 6+6 layers; the inner feedforward layer contains 2048 cells. The adapter-based systems (see below) additionally use an adaptation block in each layer, composed of a 2-layer perceptron, with an inner $\operatorname{ReLU}$ activation function operating on normalized entries of dimension~1024. 
Training uses batches of~12,288 tokens, Adam with parameters $\beta_1=0.9$, $\beta_2= 0.98$, Noam decay ($warmup\_steps=4000$), and a dropout rate of $0.1$ in all layers.

@@

\subsection{Fine tuning experiments}

\fyTodo{Fix the table}
In Table~\ref{tab:finetuning}, we compare the results of Fine tuning a system initially trained with a uniform mixture of domains with the result of learning from scratch, also with a initial uniform distribution ($\phi_0(d) = 0 \forall d$). \fyTodo{What do we see ? How about domain proximity ? - show a graph}. Note that the DDS systems are somewhat quicker to build, as we only train for 100K iterations (instead of 100K + 50K for the fine tuning part).

@@

\begin{table*}[htb]
  \centering% \small
  \begin{adjustbox}{width=1.0\textwidth,center}
  \begin{tabular}{|p{3.0cm}|*{13}{r|}} \hline
    \multirow{2}{*}{Name} & \multirow{2}{*}{$sim\_iter$} & \multirow{2}{*}{parameterization} & \multirow{2}{*}{entropy constraint} & \multicolumn{6}{|c|}{BLEU} & \multirow{2}{*}{BLEU average} \\ \cline{5-10}	
   & & & & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{it}} & \multicolumn{1}{c|}{\domain{ rel }} & \multicolumn{1}{c|}{\domain{ talk}} &  \\
    \hline
  \system{$DDS\_FT (med)$} & 1 & softmax & no & 36.7&51.14&52&44.32&90.41&33.22&51.3\\
  \system{$DDS\_FT\_law$} & 1 & softmax & no &36.05&56.18&53.57&44.05&91.24&33.09&52.36\\
  \system{$DDS\_FT\_bank$} & 1 & softmax & no &36.15&54.17&54.29&41.33&89.95&31.54&51.24 \\
  \system{$DDS\_FT\_IT$} & 1 & softmax & no &34.39&48.98&52.82&46.8&85.3&31.37&49.94 \\
  \system{$DDS\_FT\_rel$} & 1 & softmax & no & 34.54&52.29&51.46&44.8&91.77&31.84&51.12\\
  \system{$DDS\_FT\_talk$} & 1 & softmax & no & 35.53&50.55&52.64&44.86&85.8&33.47&50.48\\
  \system{$DDS\_FT\_full$} & NA & NA & NA & 37.74&59.21	&54.49&46.81&90.77&33.98&53.83\\
    \hline
    $\Delta$ & \\
  \end{tabular}
  \end{adjustbox}
  \caption{DDS as a fine-tuning receipe. We report scores with DDS and each target domain, and compare with full-fine-tuning. Note that the last line reports the results of 6 different fine tuning procedures.}
  \label{tab:finetuning}
\end{table*}
\fyTodo{Works best for small domains ?}
\fyTodo{Curves for bank / medical ou law}

@@

\subsection{Multi-domain experiments}
Results for our multidomain experiments are in Table~\ref{tab:multidomains}.

@@

Looking again at the evolution of $\lambda^l$ during training we often observe a rich get richer effect, where a small number of domains manage to attract most of the probability mass, while other are completely ignore after a given number of iterations.

\fyTodo{Explain issues}

>>>>

\newcommand{\fyTodo}[1]{\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyTodostar}[1]{\Todo*[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyDone}[1]{\done[FY]\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyFuture}[1]{\done[FY]\Todo[FY:]{\textcolor{red}{#1}}}
\newcommand{\fyDonestar}[1]{\done[FY]\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\revision}[1]{\textcolor{red}{#1}}
\newcommand{\revisiondel}[1]{}
\newcommand{\src}{\ensuremath{\mathbf{f}}} % source sentence
\newcommand{\trg}{\ensuremath{\mathbf{e}}} % target sentence
\newcommand{\domain}[1]{\texttt{\textsc{#1}}}
\newcommand{\system}[1]{\texttt{{#1}}}

\newcommand{\vlambda}{\ensuremath{\boldsymbol\lambda}\xspace} % parameters vector for a distribution
\newcommand{\vtheta}{\ensuremath{\boldsymbol\theta}\xspace} % parameters vector for a distribution
\newcommand{\vpsi}{\ensuremath{\boldsymbol\psi}\xspace} % parameters vector for a distribution
\newcommand{\indic}[1]{\ensuremath{\mathbb{I}(#1)}}
% \newcommand{\SB}[1]{\textcolor{green}{#1}}
% \newcommand{\SW}[1]{\textcolor{red}{#1}}
\newcommand{\SB}[1]{\textbf{#1}}
\newcommand{\SW}[1]{\underline{#1}}
% limits underneath
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\renewcommand\textfraction{.1}
\renewcommand\floatpagefraction{.95}
\newcommand{\sbcl}[2]{{\scriptsize #1 \hfill $|$ \hfill  #2}}
\usepackage{multirow}
\usepackage{adjustbox}

@@

\begin{document}
\maketitle
\begin{abstract}
  When building machine translation systems, one often needs to make the best out of heterogeneous sets of parallel data in training to achieve the best test performance in one or several domain(s) of interest. This multi-source / multi-domain adaptation problem is often approached with instance selection or reweighting strategies, most of which pre-suppose an ex-ante assessment of the relevance of training instances. In this paper, we study the recently proposed Differential Data Selection (DDS) model and explore its ability to serve as a generic framework for these various situations. Our experiments study both domain adaptation and multi-domain learning and show that DDS often enables to outperform more heuristic adaptation strategies. We also introduce variants that boosts DDS performance of adapter-based multi-domain systems.\fyTodo{Keep this part ?}
%   This is a well-known issue, which has given raise to a rich litterature in under the umbrellas of domain adaptation or multi-domain

@@

\section{Introduction}\label{sec:intro}
A typical setting in machine translation (MT) is to collect the largest possible collection of parallel data for the chosen language pair, with the intent to achieve optimal performance for a task of interest. In such situations, the training data distribution is opportunistic, while the test data distribution is chosen and fixed; a key component in training is then to mitigate the detrimental effects of a possible mismatch between these distributions. Single-source and multi-source domain adaptation (DA) is a well-studied instance of this setting (see \cite{Chu2017comparison} for a review) , and so is multi-domain (MD) learning \cite{Chu18multilingual,Zeng18multidomain,Jiang19multidomain,Pham21revisiting}. A related situation is multilingual machine translation \cite{Firat16multiway,Ha16towards,Johnson17google,Arivazhagan19massively}\fyTodo{Add more recent work}, where the heterogeneity of training data not only corresponds to variations in topic, genre or register, but also in language.

@@

It is usual in MT to opportunistically collect training samples from several domains, which means that training instances are distributed according to the mixture $\mathcal{D}^s$ such that $\mathcal{D}^s(x) = \sum_{d=1}^{n_d} \lambda^{s}(d) \mathcal{D}_d(x)$, where $\{\lambda^{s}(d), d=1 \dots n_d\}$ are the corresponding mixture weights satisfying $\sum_d \lambda^{s}(d)=1$.

The main challenge is then to make the best of this heterogeneous data, with the aim to achieve the optimal performance for the intended test conditions. These might correspond to data from just one of the training domains, as in standard supervised (multi-source) domain adaptation; a more difficult case is when the test data is from one domain unseen in training (unsupervised domain adaptation); in multi-domain adaptation finally the test distribution is itself a mixture of domains, some of which may also be observed in training.  Without loss of generality, one may then assume that the test distribution takes the form $\mathcal{D}^{t}(x) = \sum_d \lambda^{t}(d) \mathcal{D}_d(x)$ - with only one non-null  component in the case of fine-tuning.
These situations are illustrated in Figure~\ref{fig:mdmt-lambdas}.
\begin{figure}[h]
  \centering
  \includegraphics[width=0.48\textwidth]{mdmt-lambdas}
  \caption{Training and testing with distribution mismatch. We consider just three domains, and represent vectors of mixture weights $\vlambda^{s}$ and $\vlambda^{t}$ in the 3-dimensional simplex. Training with weights in (a) and testing with weights in (c) is supervised multi-source domain adaptation to domain~2 ($d_2$), while (b)-(c) is the unsupervised version, with no training data from $d_2$; training with weights in (a) and testing with weights in (d) is multi-domain learning, also illustrated with configurations (a)-(e) (training domain $d_1$ is not seen in test), and (b)-(d)  (test domain $d_2$ is unseen in training).}\label{fig:mdmt-lambdas}
\end{figure}

@@

A more practical standard approach to supervised DA is \emph{fine-tuning} \cite{Luong15stanford,Freitag16fast}, where $\vlambda^{l}$ is allowed to vary during the course of learning. With our notations, this approach amounts to first learning an initial parameter value with all the data ($\forall d, \lambda^{l}(d) = \lambda^{s}(d)$), then to continue training with only batches from the test domain $d_t$ ($\lambda^{l}(d) = \indic{d = d_t}$) with $\indic{A}$ the indicator function for predicate $A$. Note that this strategy is potentially suboptimal, as some out-of-domain samples may contribute to the final performance due to eg.\ domain overlap. Optimizing the learning distribution in multi-domain settings is even more challenging, as the learner has to best take advantage of potential domains overlaps, and also of the fact that some domains might be easier to learn than others.\fyTodo{How to measure this?}

Differentiable data selection (DDS), proposed by \citet{Wang20optimizing} and later applied to multilingual machine translation by \citet{Wang20balancing}, introduces an additional degree of freedom and allows to iteratively learn an optimal value for $\lambda^{l}$, as we now explain.

@@

\section{Differential Data Selection} \label{sec:dds}
\subsection{Basic principles}
Assuming training data from $n_d$ domains $d_1 \dots d_{n_d}$, we denote the size of the training corpus from domain $d$ as  $N^{s}_d$, and $N^{s} = \sum_d N^{s}_d$ is the total number of training samples. We use $\widehat{\mathcal{D}^l_d}$ and $\widehat{\mathcal{D}^t _d}$ to denote the empirical train and test distributions for domain $d$, and \revision{$\widehat{\mathcal{D}^{u}}(x;\lambda^{u}) = \sum_{d} \lambda^{u}(d) \widehat{\mathcal{D}^{u}_d}(x)$ for $u\in\{l,t\}$} \fyTodo{$\widehat{\mathcal{D}}(x;\lambda) = \sum_{d} \lambda(d) \widehat{\mathcal{D}_d}(x)$}. In our setting,  $\vlambda^t$, and hence $\widehat{\mathcal{D}^t}(x;\vlambda^t)$ are fixed and predefined, approximated with an equivalent number of development corpora. Training optimizes the parameters $\vtheta$ of a neural translation model.

Sampling training domains using a distribution $\lambda^{l}$ results in a parameter estimate $\theta(\lambda^{l})$, delivering performance that can be estimated on the development set using $\lambda^{t}$. \citet{Wang20optimizing,Wang20balancing} then formulate the problem of finding an optimal sampling strategy $\lambda^{l}_{*}$ maximizing performance on the development data as the following bi-level optimization problem:
\begin{equation} \label{eq:bilevel}
\begin{split}
\vlambda^{l}_* &= \displaystyle{\mathop{\argmin}_{\lambda^{l}} J(\theta^*(\vlambda^{l}), \vlambda^{t})} \\
					& \theta^*(\vlambda^{l}) = \displaystyle{\mathop{\argmin_{\vtheta}} J(\vtheta, \vlambda^{l})}, 
\end{split}
\end{equation}
where we use the following notations:% \fyTodo{Overload Q, from domain to instances to pairs}

\begin{align*}
J(\vtheta, \vlambda^{t} ) &= \displaystyle{\mathop{\sum}_{(x,y) \sim \widehat{\mathcal{D}^{t}(x)}} \ell(\vtheta,x,y)} \\
J(\vtheta, \vlambda^{l}) &= \displaystyle{\mathop{E_{(x,y) \sim \widehat{\mathcal{D}^{l}(x)}}[\ell(\vtheta,x,y)]}} \\
l(\vtheta,x,y) &= - log P(x,y;\vtheta)
\end{align*}
Problem~\eqref{eq:bilevel} can be approached by coordinate gradient descent alternating between $\lambda^{l}$ and $\vtheta$. \citet{Wang20balancing} limit the search space of $\vlambda^{l}$ to a parameterized differentiable function $\vlambda^{l}(\vpsi)$. Given $\vpsi_t$ at time step $t$, $\vtheta_t$ is updated via usual gradient descent as follows:
\begin{align*}
\vtheta_{t+1} &= \vtheta_t - \mathbf{lr}_{nmt} * \frac{\partial l(\vtheta_t, x,y)}{\partial \vtheta} \\
x,y &\sim \widehat{\mathcal{D}^l}(x;\vlambda^l(\psi_t)) = \displaystyle{\mathop{\sum}_{d=1}^K} \lambda^l(d;\psi_t) * \widehat{D_d}(x) \\
\end{align*}
Given $\vtheta_t$, $\psi_t$ is updated via the Reinforce algorithm \cite{Williams92reinforce}:\fyTodo{Rewrite this correctly with domain average reward}
%\begin{align*}
%  \psi_{t+1} &= \psi_t + \mathbf{lr}_{data} * R(x,y) * \\
%  & \quad\quad \frac{\partial log(\lambda^l(\psi_t)(x,y))}{\partial \psi} \\
%x,y &\sim \lambda^l (\psi_t) \\
%R(x,y) &= \langle \frac{\partial l(\theta',x,y)}{\partial \theta}, \frac{\partial J(\theta_{t+1}, \lambda^l)}{\partial \theta} \rangle \\
%\theta' &= \theta_t - \mathbf{lr}_{nmt} * \frac{\partial l(\theta_t, x,y)}{\partial \theta} \\
%\end{align*}

@@

This setting is pretty general and can in principle accomodate the variety of situations mentioned  above, and many more: basic domain adaption, multi-domain adaptation with various target distributions, possibly including domains unseen in training. In our experiments we would like to better assess the actual potential of DDS is these settings and seek to experimentally answer the following questions:
\begin{itemize}
\item is DDS a viable alternative to conventional fine-tuning? In particular does it enable to better take advantage of relevant data from other domains?
\item is DDS also a viable option in multidomain adaptation scenarios?
\item does DDS also enable to perform unsupervised (multi-) domain adaptation?\fyTodo{TBContinued}
\end{itemize}

@@

\subsection{Data and metrics \label{ssec:corpora}}
We experiment with translation from English into French and use texts initially originating from 6~domains, corresponding to the following data sources: the UFAL Medical corpus V1.0 (\domain{med})\footnote{\url{https://ufal.mff.cuni.cz/ufal_medical_corpus}.\revision{We only use the in-domain (medical) subcorpora: PATR, EMEA, CESTA, ECDC.}}; the European Central Bank corpus (\domain{bank}) \cite{Tiedemann12parallel}; the JRC-Acquis Communautaire corpus (\domain{law}) \cite{Steinberger06acquis}; documentations for KDE, Ubuntu, GNOME and PHP from the Opus collection \cite{Tiedemann09news}, merged in a \domain{it}-domain; TedTalks (\domain{talk}) \cite{Cettolo12wit}, and the Koran (\domain{rel}). Complementary experiments also use v12 of the News Commentary corpus (\domain{news}). Most corpora are available from the Opus website.\footnote{\url{http://opus.nlpl.eu}} These corpora were deduplicated and tokenized with in-house tools; statistics are in Table~\ref{tab:Corpora}. To reduce the number of types and build open-vocabulary systems, we use Byte-Pair Encoding \cite{Sennrich16BPE} with 30,000 merge operations on a corpus containing all sentences in both languages.\fyDone{Add \# number of tokens, also specificity ?}%

@@

\begin{itemize}
\itemsep0em 
\item Generic models trained with various predefined mixtures of the training data taking the form:
\begin{align} \label{mixture:trn}
\lambda^{l}(d;\alpha) = \frac{q_d^{\alpha}}{\displaystyle{\mathop{\sum}_{d=1}^{n_d}q_d^{\alpha}}} &&
q_d = \frac{\mid N^{s}_d \mid}{\displaystyle{N^{s}}} % \mathop{\sum}_{i=1}^K\mid D_i \mid}}
\end{align} 
with $\alpha \in [0,0.25,0.5,0.75,1.0]$. These systems are denoted \system{Mixed-$\alpha$} below. \system{Mixed-$0$} uses a uniform domain distribution, \system{Mixed-$1.0$} simple uses the observed domain distribution.
%on a concatenation of all corpora (\texttt{Mixed}). We develop two versions\footnote{In fact three: to enable a fair comparison with WDCMT, a RNN-based variant is also trained and evaluated. \revision{This system appears as \system{Mixed-Nat-RNN} in Table~\ref{tab:performance}}.} of this system, one where the domain unbalance reflects the distribution of our training data \revision{given in Table~\ref{tab:Corpora}} (\system{Mixed-Nat}) and one where all domains are equally represented in training (\system{Mixed-Bal}). The former is the best option when the train mixture $\mathcal{D}^s$ is also expected in testing; the latter should be used when the test distribution is uniform across domains. Accordingly, we report two aggregate scores: a weighted average reflecting the training distribution, and an unweighted average, meaning that test domains are equally important.
\item fine-tuned models \cite{Luong15stanford,Freitag16fast}, based on the \system{Mixed-$1.0$} system, further trained on each domain for at most 20~000 iterations, with early stopping when the dev BLEU stops increasing. The full fine-tuning (\system{FT-Full}) procedure may update all the parameters of the initial generic model, resulting in six systems each adapted for one domain, with no parameter sharing across domains.

@@

All models use embeddings and hidden layers sizes of dimension~512. Transformer models contain 8~attention heads in each of the 6+6 layers; the inner feedforward layer contains 2048 cells. The adapter-based systems (see below) additionally use an adaptation block in each layer, composed of a 2-layer perceptron, with an inner $\operatorname{ReLU}$ activation function operating on normalized entries of dimension~1024. 
Training uses batches of~12,288 tokens, Adam with parameters $\beta_1=0.9$, $\beta_2= 0.98$, Noam decay ($warmup\_steps=4000$), and a dropout rate of $0.1$ in all layers.

@@

\subsubsection{Fine-tuning experiments}

\fyTodo{Fix the table}
In Table~\ref{tab:finetuning}, we compare the results of fine-tuning a system initially trained with a uniform mixture\fyTodo{uniform or natural?} of domains with the result of learning from scratch, also with a initial uniform distribution ($\phi_0(d) = 0 \forall d$). \fyTodo{What do we see ? How about domain proximity ? - show a graph}. Note that the DDS systems are somewhat quicker to build, as we only train for 100K iterations (instead of 100K + 50K for the fine-tuning part).

@@

\begin{table*}[htb]
  \centering% \small
  \begin{adjustbox}{width=1.0\textwidth,center}
  \begin{tabular}{|p{3.0cm}|*{13}{r|}} \hline
    \multirow{2}{*}{Name} & \multirow{2}{*}{$sim\_iter$} & \multirow{2}{*}{parameterization} & \multirow{2}{*}{entropy constraint} & \multicolumn{7}{|c|}{BLEU} \\ \cline{5-11}	
   & & & & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{it}} & \multicolumn{1}{c|}{\domain{ rel }} & \multicolumn{1}{c|}{\domain{ talk}} & \multicolumn{1}{c|}{mean} \\
    \hline
  \system{$DDS\_FT (med)$} & 1 & softmax & no & 36.7&51.14&52&44.32&90.41&33.22&51.3\\
  \system{$DDS\_FT\_law$} & 1 & softmax & no &36.05&56.18&53.57&44.05&91.24&33.09&52.36\\
  \system{$DDS\_FT\_bank$} & 1 & softmax & no &36.15&54.17&54.29&41.33&89.95&31.54&51.24 \\
  \system{$DDS\_FT\_IT$} & 1 & softmax & no &34.39&48.98&52.82&46.8&85.3&31.37&49.94 \\
  \system{$DDS\_FT\_rel$} & 1 & softmax & no & 34.54&52.29&51.46&44.8&91.77&31.84&51.12\\
  \system{$DDS\_FT\_talk$} & 1 & softmax & no & 35.53&50.55&52.64&44.86&85.8&33.47&50.48\\
  \system{$FT\_full$} & NA & NA & NA & 37.74&59.21	&54.49&46.81&90.77&33.98&53.83\\
    \hline
    $\Delta$ & \\
  \end{tabular}
  \end{adjustbox}
  \caption{DDS as a fine-tuning receipe. We report scores with DDS and each target domain, and compare with full-fine-tuning. Note that the last line reports the results of 6 different fine tuning procedures.}
  \label{tab:finetuning}
\end{table*}

\fyTodo{This table does not work- One line would be enough, where we compare autoFT with true FT}

\fyTodo{Works best for small domains ?}
\fyTodo{Curves for bank / medical ou law}

@@

\subsubsection{Multi-domain experiments}
Results for our multidomain experiments are in Table~\ref{tab:multidomains}.\fyTodo{Same table, line autoMD. We should be clear about the dev / test distribution - uniform ? so I would compare MIX with DDS(UNIF, UNIF) - DDS(NAT,UNIF) and not report / domain scores}

@@

In this scenario, DDS is not as good as using a fixed distribution, with a well-chosen $\alpha$.
